\chapter{Results \& Evaluation}
\label{sec:results}

Previously, formalisms and proofs of correctness for two extant
garbage collection algorithms were developed, and from this a more
generalised notion of correctness was extracted. This chapter
discusses and evaluates the results of the two facets of the work, and
how successfully the goals of the project have been met. An
experimental validation of the proofs involving an implementation of
the two collectors is also be discussed.

\section{Proofs of Collectors}
\label{sec:results-collectors}

In the process of producing a generalised notion of correctness for
garbage collectors, proofs of the Armstrong/Virding and
Fenichel/Yochelson collectors were first provided: two very different
approaches, but both for systems based on fixed-size cons cells. The
proofs for neither collector could be machine verified in their
current form, however they remain convincing despite this.

The collectors were chosen for their simplicity: they are both old,
established, collectors which have stood the test of time, and they
are sufficiently simple that one can `see' their correctness at a
glance. They do not attempt to perform any nontrivial cleverness to
improve the performance of the mutator or memory system, as more
modern hybrid collectors are wont to do, and they fall strictly within
the bounds of mark-sweep and copying collectors, without borrowing
anything from other areas. Thus, the proofs simply make explicit the
intuition we feel when we read them, and the implementation of the
collectors provides an experimental validation of this intuition.

The proofs may appear quite different on the surface, however many of
the definitions used in the mark-sweep proof were reused in the
copying proof, and those that were changed were made more
general. The proofs may not be so different after all,
then. Naturally, there are some parts which are collector-specific
(such as correct marking, sweeping, and copying), but these arose from
trying to convert the very abstract postcondition into something more
amenable to proof.

\section{Collector Implementation}
\label{sec:results-impl}

As evidence of the correctness of the proofs, both of the collectors
were implemented\footnote{All of the code produced can be found in
  Appendix \ref{sec:gc-impl}, and online at
  \url{https://github.com/barrucadu/meng-project}.} in C, with a small
test program with a very constrained heap which constructs a variety
of linked lists, and results in over 200 garbage collections
occurring. Assertions were then attached to points in the programs to
verify the preconditions, postconditions, and invariants of the
collectors.

The implementation does have some limitations imposed upon it for
simplicity: there is a fixed-size array of roots, which the programmer
must ensure contains all roots, and the requirement of immutable cells
in the Armstrong/Virding collector is not implemented. Resolving both
of these would have been significantly more work, and is not necessary
in order to demonstrate that the garbage collection works.

Both collectors manage a heap of the same number of cells (in the case
of Fenichel/Yochelson, both semispaces are this big), in order to have
identical characteristics, allowing identical assertions to be used in
the test program. As neither collector comes with an allocator,
allocators have been produced which work with them. The way that this
has been done is not necessarily the only, or best, way, it is just
what appears to work. In both cases, the collector is invoked by the
allocator upon not having enough memory left for an allocation. The
preconditions are checked before calling the garbage collection
function, and the postconditions checked after. In each of the garbage
collector loops, the invariants are checked after each iteration.

Due to using a low level language, the implementations of the
collectors look somewhat more complex than what is specified in their
respective papers, but is as close as could reasonably be obtained.

The assertions, and their implementation, are discussed below.

\subsection{Armstrong/Virding}
\label{sec:results-impl-ms}

\subsubsection{Preconditions}
\label{sec:results-impl-ms-pre}

The Armstrong/Virding collector has a precondition that the
\texttt{first} cell is always reachable, as otherwise it would be
garbage, and due to it never being freed, garbage would survive a
collection. Furthermore, it is assumed that when the collector is
called, all cells pointed to by roots have been marked.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
assert(reachable(gchead(first)));

for(unsigned int i = 0; i < NUM_ROOTS; i++)
  if(roots[i] != NULL)
    gchead(roots[i])->mark = MARKED;
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Armstrong/Virding Preconditions}
\end{figure}

\subsubsection{Postconditions}
\label{sec:results-impl-ms-post}

The postconditions express simply the correct mark-sweep
postcondition: after collection, all cells which are allocated (not on
the free list) are reachable from the roots. Furthermore, the live
cell invariant is checked here as a postcondition, although arguably
it should be checked with the rest of the invariants, in the garbage
collection loop.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&heap[i]))
    assert(!on_free_list(&heap[i]));

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(!on_free_list(&heap[i]))
    assert(reachable(&heap[i]));
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Armstrong/Virding Postconditions}
\end{figure}

\subsubsection{Invariants}
\label{sec:results-impl-ms-invariants}

The invariants appear somewhat more complex. The first loop expresses
$\forall c \in h,\ \lnot\free{c} \land \id~c > \id~\mathtt{SCAV} \implies
(\forall c \pointsto x, \id~x < \id~\mathtt{SCAV} \implies
\marked{h}{c})$: all cells outside of the region that has been
collected, but which are pointed to by a cell in that region, have
been marked. The latter two express the simpler properties that all
allocated cells in the collected region are live, and all allocated
cells in the collected region are unmarked.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&heap[i]) && cell_id(&heap[i]) > cell_id(gchead(SCAV)))
    {
      if(heap[i].cell.car.tag == REFERENCE &&
         heap[i].cell.car.val.ptr != NULL &&
         cell_id(gchead(heap[i].cell.car.val.ptr)) < cell_id(gchead(SCAV)))
        assert(gchead(heap[i].cell.car.val.ptr)->mark == MARKED);

      if(heap[i].cell.cdr.tag == REFERENCE &&
         heap[i].cell.cdr.val.ptr != NULL &&
         cell_id(gchead(heap[i].cell.cdr.val.ptr)) < cell_id(gchead(SCAV)))
        assert(gchead(heap[i].cell.cdr.val.ptr)->mark == MARKED);
    }

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&heap[i]) && cell_id(&heap[i]) >= cell_id(gchead(SCAV)))
    assert(reachable(&heap[i]));

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&heap[i]) && cell_id(&heap[i]) > cell_id(gchead(SCAV)))
    assert(heap[i].mark == UNMARKED);
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Armstrong/Virding Invariants}
\end{figure}

\subsection{Fenichel/Yochelson}
\label{sec:results-impl-c}

\subsubsection{Preconditions}
\label{sec:results-impl-c-pre}

An invariant which was not mentioned whilst discussing the proof, but
which the garbage collector, mutator, and allocator must obey, is that
there are no inter-semispace pointers. Due to a bug in the test
mutator code (which has since been remedied), a garbage collection
occuring whilst appending a value to a list would result in this
invariant being broken, and resulted in some confusion before the
issue became apparent: this was then enforced as a precondition of the
collector.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
cell *sspace = (cons_space == FIRST) ? semispace1 : semispace2;

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&sspace[i]))
    {
      if(sspace[i].car.tag == REFERENCE &&
         sspace[i].car.val.ptr != NULL)
        assert(sspace[i].car.val.ptr >= sspace &&
               sspace[i].car.val.ptr <= &sspace[NUM_CELLS]);

      if(sspace[i].cdr.tag == REFERENCE &&
         sspace[i].cdr.val.ptr != NULL)
        assert(sspace[i].cdr.val.ptr >= sspace &&
               sspace[i].cdr.val.ptr <= &sspace[NUM_CELLS]);
    }
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Fenichel/Yochelson Preconditions}
\end{figure}

\subsubsection{Postconditions}
\label{sec:results-impl-c-post}

As a postcondition the live cell invariant was again checked, and also
it was asserted that all allocated cells are reachable.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
cell* sspace = (cons_space == FIRST) ? semispace1 : semispace2;

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(reachable(&sspace[i]))
    assert(!on_free_list(&sspace[i]));

for(unsigned int i = 0; i < NUM_CELLS; i++)
  if(!on_free_list(&sspace[i]))
    assert(reachable(&sspace[i]));
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Fenichel/Yochelson Postconditions}
\end{figure}

\subsubsection{Invariants}
\label{sec:results-impl-c-invariants}

The invariant chosen to be verified is that, after collecting every
root, all pointers in cells reachable from that root have been
updated; however, this proved too strong to implement concisely, and
so the weaker property that pointers pointed into the correct
semispace was verified. Unfortunately, proving the stronger property
would require having access to the pre-state of the heap, which would
require copying it before initiating a garbage collection.

\begin{figure}[H]
  \begin{lstlisting}[language=C]
for(unsigned int i = 0; i < NUM_ROOTS; i ++)
  if(roots[i] != NULL)
    {
      roots[i] = collect(roots[i]);
      check_pointers_updated(roots[i], NULL);
    }
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Fenichel/Yochelson GC Loop}
\end{figure}

\begin{figure}[H]
  \begin{lstlisting}[language=C]
static void check_pointers_updated(const cell* root, const cell* cur)
{
  if(root == cur)
    return;

  if(cur == NULL)
    cur = root;

  cell *sspace = (cons_space == FIRST) ? semispace1 : semispace2;

  if(cur->car.tag == REFERENCE &&
     cur->car.val.ptr != ALREADYCOPIED &&
     cur->car.val.ptr != NULL)
    {
      assert(cur->car.val.ptr >= sspace &&
             cur->car.val.ptr <= &sspace[NUM_CELLS]);
      check_pointers_updated(root, cur->car.val.ptr);
    }

  if(cur->cdr.tag == REFERENCE &&
     cur->cdr.val.ptr != NULL)
    {
      assert(cur->cdr.val.ptr >= sspace &&
             cur->cdr.val.ptr <= &sspace[NUM_CELLS]);
      check_pointers_updated(root, cur->cdr.val.ptr);
    }
}
  \end{lstlisting}
  \captionsetup{format=default}
  \caption{Fenichel/Yochelson Pointer Checking}
\end{figure}

\section{Generalised Garbage Collection Correctness}
\label{sec:results-correctness}

Having produced two proofs of correctness, it was then attempted to
extract a more general formalism, however, it became apparent that the
formalism for copying collectors was general enough to cover both of
the cases already. Rather than conclude that the project was done,
generalisation was pursued in a different direction, by allowing the
garbage collector to mutate metadata in cells.

Whilst initially appearing to only be a small change, this actually
allows a much larger class of garbage collectors to be reasoned about,
and an example is provided below. A general garbage collector was then
split up into a series of abstract stages, which can be reasoned about
separately, and which aids proof. It was also shown how these could be
specialised to the cases of the mark-sweep and copying collectors
already considered.

\begin{example}
  Consider a simple generational garbage collector, which tracks what
  generation a cell belongs to by giving it an integer tag, numbered 0
  for the younger generation, 1 for the next youngest, and so on, up
  to some oldest generation $n$.

  If we can mutate the metadata, we can promote a cell to the next
  generation by incrementing this tag in the collector, whereas in the
  copying formalism we would be unable to do this, as only pointers
  can be mutated. This would have to happen outside of the garbage
  collector, in some separate `generation incrementer', but
  separating these two things does not appear to make much sense, and
  is just an inconvenience imposed by an overly-strict formalism.
\end{example}

We can see here that, whilst the intent was to produce a general
formalism by showing that the mark-sweep and copying collectors were
rather different (but still sharing a common theme), this
generalisation is still worthwhile and interesting, as it allows more
collectors to be reasoned about than perhaps would have been possible
if the initial goal had been pursued.

\section{Summary}
\label{sec:results-summary}

To summarise the results, proofs of correctness for two garbage
collectors were produced, in the process formalising everything needed
to achieve this. The formalisms were then generalised to cover an even
greater range of collectors. Finally, the two collectors were
implemented, with pre- and post-conditions checked throughout.

The proofs were discussed in Section \ref{sec:results-collectors},
with some further comments on the proofs arising from the discussion
of the implementations in Section \ref{sec:results-impl}. Finally, a
motivating example was given to illustrate the utility of the more
general formalism in Section \ref{sec:results-correctness}.
