\documentclass[10pt,a4paper,twoside,openright]{report}

% Input
\usepackage[utf8x]{inputenc}

% Nicer fonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% Graphics support
\usepackage{graphicx}
\usepackage{epstopdf}
\DeclareGraphicsRule{.eps}{pdf}{.pdf}{`epstopdf #1}

% Hyperlinks, URLs etc.
\usepackage{hyperref}
\usepackage{url}
\hypersetup{
    colorlinks=true,
    citecolor=black,
    urlcolor=black,
    linkcolor=black,
    pagecolor=black,
    anchorcolor=black
}

% Stuff at the end
\usepackage[toc,page]{appendix}
\usepackage[xindy]{glossaries}
\usepackage[xindy]{imakeidx}
\loadglsentries{glossary.glo}
\renewcommand*{\glossaryentrynumbers}[1]{}
\makeglossaries
\makeindex

% Advanced maths features
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
% Conditional macros
\usepackage{xifthen}

% Definitions and theorems
\usepackage{amsthm}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

% Drawings
\usepackage[usenames,dvipsnames]{color}
\usepackage{tikz}
\usetikzlibrary{trees}

% Red underlined TODO messages!
\usepackage[normalem]{ulem}
\newcommand{\todo}[1]{{\color{red}{\uline{#1}}}}
% [citation needed]
\newcommand{\cn}{{\color{red}$^{\text{[citation needed]}}$}}

% Page layout jiggery pokery
\usepackage{pdflscape}
\usepackage{multicol}

\newenvironment{wide}{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \addtolength{\leftmargin}{-1.5cm}%
      \addtolength{\rightmargin}{-1.5cm}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]%
}{%
  \end{list}%
}

\newenvironment{lscape}{\begin{landscape}\begin{wide}}{\end{wide}\end{landscape}}

% Logic & Maths
\usepackage{bussproofs}

\newcommand{\htriple}[3]{\left\{#1\right\}\ \mbox{#2}\ \left\{#3\right\}}
\newcommand{\len}{\mathrm{len}~}
\newcommand{\true}{\mathbf{true}}
\newcommand{\false}{\mathbf{false}}
\renewcommand{\iff}{\Leftrightarrow}
\renewcommand{\implies}{\Rightarrow}

\newcommand{\finparmap}{\rightharpoonup}
\newcommand{\pointsto}{\hookrightarrow}
\newcommand{\restrict}{\downharpoonright}
\newcommand{\addr}{\mathrm{Addr}}
\newcommand{\data}{\mathrm{Data}}
\newcommand{\nullp}{\mathit{null}}
\newcommand{\listof}[1]{#1\ \mathbf{list}}
\newcommand{\x}{\times}
\newcommand{\of}{\circ}
\newcommand{\reach}[2]{\mathrm{reach}\left(#1, #2\right)}
\newcommand{\garbage}[2]{\mathrm{garbage}\left(#1, #2\right)}
\newcommand{\keep}[2]{\mathrm{keep}\left(#1, #2\right)}
\newcommand{\markset}[2]{\mathrm{mark}\left(#1, #2\right)}
\newcommand{\map}[2]{\mathrm{map}\left(#1, #2\right)}
\newcommand{\rename}[2]{\mathrm{rename}\left(#1, #2\right)}
\newcommand{\gc}{\xrightarrow{gc}}
\newcommand{\translate}{\xrightarrow{translate}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\dom}{\mathrm{dom}}
\newcommand{\gchead}[1]{\mathrm{gchead}\left(#1\right)}
\newcommand{\alloc}[1]{\mathrm{alloc}\left(#1\right)}
\newcommand{\marked}[2]{\mathrm{marked}\left(#1, #2\right)}

% Listings
\usepackage{listings}

\definecolor{LightGray}{rgb}{0.9,0.9,0.9}

\lstset{basicstyle=\small\ttfamily}
\lstset{showstringspaces=false}
\lstset{numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt}
\lstset{keywordstyle=\color{MidnightBlue}\bfseries}
\lstset{commentstyle=\color{JungleGreen}}
\lstset{identifierstyle=\color{OliveGreen}}
\lstset{stringstyle=\color{Red}}
\lstset{backgroundcolor=\color{LightGray}}
\lstset{breaklines=true}

\title{%
{\normalsize\scshape University of York}\\
\vspace{2em}
\hrule
\vspace{1em}
{\huge\bfseries Memory Management}\\
\vspace{1em}
{\bfseries Verified Garbage Collection}\\
\vspace{1em}
\hrule
\vspace{1em}
{\small Submitted for part completion of MEng Computer Systems and Software Engineering}
\vfill%
{\normalsize Number of words: TODO as counted by ``detex | wc -w'',\\
excluding code listings and appendices.}
}

\author{%
Author: Michael Walker\\
Supervisor: Colin Runciman
}

\date{}

\begin{document}

\pagestyle{empty}

\maketitle

\cleardoublepage
\begin{abstract}
\todo{Not a mini introduction. Write better abstract after having done
  more work}

Programming languages with managed runtimes abstract away from much,
sometimes even all, of the need for manual memory management. Whilst
a lot of work has gone into how to make these critical systems
performant, there has not been so much work on formal verification,
and developers often rely on traditional testing and bug
reporting. This is far from ideal, as source-level verification of
software is pointless if the runtime may have arbitrary bugs lurking
within.

In this project, I propose a formalism of partial correctness for
garbage collection, and then show how some common algorithms are
correct under this formalism. Finally, I propose a new garbage
collection algorithm based on a review of the literature, prove that
it is correct, and then measure the performance of a (non-verified)
Java implementation in Jikes RVM.  \vfill

\paragraph{Statement of Ethics}

This project concerns only the development and implementation of
algorithms. No human participation is involved at any point and as
such there are no specific ethical implications.
\end{abstract}


% Table of contents, roman numeral page numbers
\cleardoublepage
\pagestyle{plain}
\pagenumbering{roman}
\setcounter{page}{1}
\tableofcontents

%\listoffigures
%\listoftables
%\lstlistoflistings

% Reset page numbering for content
\newpage
\cleardoublepage
\pagenumbering{arabic}

\include{01-introduction}
\include{02-litreview}
\include{03-analysis}
\include{04-formalism-marksweep}
\include{05-formalism-copying}
\include{06-formalism-general}
\include{07-algorithm}
\include{08-results}
\include{09-conclusion}

\begin{appendices}
  % \addcontentsline{toc}{subsection}{Index}
  % \printindex

  \addcontentsline{toc}{subsection}{Glossary}
  \glsaddall
  \printglossaries

  \addcontentsline{toc}{subsection}{Bibliography}
  \bibliographystyle{plain}
  \bibliography{references}
\end{appendices}

\end{document}
